---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# garry

<!-- badges: start -->
<!-- badges: end -->

garry is a lazy, spatial-aware raster engine for R. You describe a whole
raster computation as a graph and it runs it fast: chunked, distributed across
processes, with the numeric kernels JIT-compiled to XLA (CPU or GPU). It is
built for cloud-native earth observation, turning a STAC search of remote COGs
into an analysis-ready composite without leaving R.

**What it is:**

- **Lazy.** Every operation (`lazy_map()`, `focal()`, `reduce_over()`, `mask()`,
  `align()`) adds a node to a computation graph and returns immediately. Nothing
  reads or computes until `collect()`.
- **Spatial-aware.** Arrays carry their CRS, transform and extent. Alignment is
  explicit: binary ops never silently resample, so a pixel is never quietly
  moved half a cell.
- **Fast.** `collect()` plans the graph into chunks, streams reads through a
  GDAL daemon pool, and runs the fused kernels through
  [anvl](https://github.com/r-xla/anvl) (XLA). On a three-band annual HLS
  composite it matches Python's ODC + dask on the same machine, in idiomatic R.
- **GDAL-faithful.** IO is [gdalraster](https://cran.r-project.org/package=gdalraster);
  grid math is [vaster](https://github.com/hypertidy/vaster). Results track what
  GDAL would give you.

## Installation

You can install the development version of garry from
[GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("belian-earth/garry")
```

## Example

A cloud-masked annual median composite of HLS Sentinel-2 imagery, straight from
the Microsoft Planetary Computer. The whole pipeline is lazy: it reads and
computes only at the final `collect()`.

```{r example, eval = TRUE}
library(garry)

# 1. Discover: HLS S30 over an AOI, all of 2023, pre-signed for the PC.
aoi <- c(144.13, -7.725, 144.47, -7.475)   # lon/lat bounding box
its <- stac_query(
  bbox        = aoi,
  stac_source = "https://planetarycomputer.microsoft.com/api/stac/v1/",
  collection  = "hls2-s30",
  start_date  = "2023-01-01", end_date = "2023-12-31"
) |>
  rstac::items_sign(rstac::sign_planetary_computer())

src <- stac_sources(its, assets = c("B04", "B03", "B02", "Fmask")) 

# 2. The analysis grid, straight from the AOI: an equal-area (LAEA) grid at 30 m
#    centred on the bbox. No hand-picked EPSG or projected extent.
target <- grid_from_bbox(aoi, res = 30)

# 3. A read/compute daemon pool, auto-sized to the machine.
garry_daemons()

# 4. Build the pipeline. A `LazyDataset` holds every band over time; the verbs
#    apply across all bands at once. Still nothing has been read or computed.
composite <- lazy_dataset(
  src, grid = target,
  assets     = c("B04", "B03", "B02"), mask_asset = "Fmask",
  nodata     = c(B04 = -9999, B03 = -9999, B02 = -9999, Fmask = 255)
) |>
  mask(from = "Fmask", where = qa_bits(0:3), open = 2, dilate = 3) |>  # clouds/shadows + cleanup
  reduce_over("median", over = "t")                                    # per-band temporal median

print(composite)

draw(composite)

draw(composite[["B04"]])

# 5. Execute across the daemons and write the GeoTIFF.
collect(composite, path = "composite.tif", nodata = -9999, distributed = TRUE)

r <- new(gdalraster::GDALRaster, "composite.tif") 
gdalraster::plot_raster(r)
r$close()
```

The same verbs work on a single raster. `lazy_source()` opens one COG or a GDAL
mosaic; `lazy_map()` / `focal()` / `reduce_over()` build map-algebra graphs;
`align()` reprojects onto a target grid; `collect()` runs it. See
`benchmarks/compare.sh` for the back-to-back garry-vs-ODC benchmark.
